---
title: Time Series Forecasting using LSTM in R
author: Richard Wanjohi, Ph.D
date: '2018-04-03'
slug: time-series-forecasting-using-lstm-in-r
categories:
  - Deep Learning
tags:
  - Keras
  - R
  - Tensorflow
output:
  blogdown::html_page:
    toc: true
    fig_width: 6
    dev: "svg"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i \alpha_i$$


$\sum_{i=1}^n X_i$

```{r chunk_name}
x <- rnorm(100)
y <- 2*x + rnorm(100)
cor(x, y)
```

There are `r nrow(iris)` individuals.

